---
title: "P8105 Homework 2"
output: github_document
date: "October 5th, 2022"
---

```{r setup, include = FALSE}
library(tidyverse)
```

## Problem 1: NYC Transit Data

### Cleaning the data

We will work with NYC Transit data; which contains information related to each entrance and exit for each subway station in NYC. 

We will read in and clean the data by retaining line, station name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. We also want to convert the entry variable from character (YES vs NO) to a logical variable. 

To achieve this, we will read in the dataset using `read_csv()`, convert the variable names to snake case using `janitor::clean_names()`, use `select()` to keep the variables of interest in our data, then use `mutate()` and `recode()` to convert the `entry` from a character to a logical variable.

```{r clean_data, warning = FALSE, message = FALSE}
transit_data = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>% 
  janitor::clean_names() %>% 
  select(line:entry, vending, ada) %>% 
  mutate(entry = recode(entry, YES = TRUE, NO = FALSE)) 

head(transit_data)
```

Using `head(transit_data)`, we can see our cleaned data includes 19 variables and 1868 observations. For each station entrance/exit, we have information about the station name, line, the station latitude and longitude, the routes it serves, entrance type, whether it has an entry, has a ticket vendor, and whether it is ADA compliant.

After these steps, these data are not fully tidy. We can see there are several `route#` columns to indicate each individual subway route a station serves, making our dataset very wide. Moreover, the `vending` variable could also be converted to a logical variable, since it is a binary variable (like `entry` and `ada`).

### Data Exploration
Next, we will explore a few aspects of our data.
Using `distinct()`, we can get the number of unique stations, accounting for stations with the same name along different lines . 

```{r}
distinct(transit_data, line, station_name) %>% 
  arrange(station_name)
```
Arranging the data by station name, we can see the output gave us entries for each unique combination of station name and line.

In total, there are **`r nrow(distinct(transit_data, line, station_name))` unique stations**.

Next, we want to know how many stations are ADA compliant. For this, we can use `group_by` and `summarise` to get the frequency of `TRUE`'s under the `ada` variable.

```{r}
transit_data %>%
  distinct(line, station_name, ada) %>% 
  group_by(ada) %>%
  summarise(n = n())
```

From our results, we can see that **84 stations are ADA-compliant.**

Finally, we want to know what proportion of entrances/exits without vending allow entrance. For this, we will `filter` observations to those with `vending == "NO"`, then use `group_by`, `summarise`, and `mutate` to compute the proportion of entrances/exits with `entry == "YES"`.

```{r}
transit_data %>%
  filter(vending == "NO") %>% 
  group_by(entry) %>%
  summarise(n = n()) %>%
  mutate(freq = n / sum(n))
```
From our results, **37.71% of station entrances / exits without vending allow entrance.**


## Problem 2: Mr. Trash Wheel Data
### Reading and cleaning data
First, we will read and clean the Mr. Trash Wheel sheet from the Trash Wheel Collection Totals Excel file in `./data`:

```{r}
trash_wheel = 
  readxl::read_excel("./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx",
                     sheet = "Mr. Trash Wheel",
                     range = "A2:N534") %>% 
  janitor::clean_names() %>% 
  filter(!is.na(dumpster)) %>% 
  mutate(sports_balls = as.integer(sports_balls))
```



## Problem 3: FiveThirtyEight Data



